Chatbots and messaging platforms in the classroom: an analysis from the teacher's perspective

### 2021-07-09 edubots_surveys.Rnw
Update edubots_surveys.Rnw

A \textit{chatbot} is a program, sometimes developed using Artificial Intelligence techniques, able to communicate in a similar way as humans do \cite{Gong2008}, using, in some cases, natural language. Indeed, most of these applications would be, apparently, close to passing a classic Turing test \cite{moor2003turing}, since they are able to answer almost any question fluently, and even pose own questions in their conversations. The recent advancement in this discipline has played an important role in many fields especially in education and online tutoring \cite{clarizia2018,Smutny2020}. These automatic systems facilitate the delivery of personalized learning by adapting to students’ pace of learning and providing customized online tutoring outside the classroom. Thus, chatbots can significantly contribute to providing interactive learning experiences as well as improving individual attention \cite{agarwal2020}.
 In this sense, chatbot technology offers a great opportunity for the improvement of tutoring systems \cite{Daniel2016,agarwal2020}, as not all students are comfortable with face-to-face tutoring with the instructor. In fact, in many cases, some students experience stress because of their need to ask a question in front of the entire class. This is why many students refrain to speak  
 % Antonio - I guess refrain to silence is the opposite of what we wanted to say, isn't it? :D
 in the classroom and tend to contact their teachers later via email.
 Consequently, this can lead to not only a delay in obtaining answers to their questions but also can cause a significant increase in their teachers’ workload, especially when factoring in the disproportion between the class size and the number of teachers. Hence, the chatbot technology has a potential to mitigate this problem by providing answers to students’ questions and facilitating a dynamic and autonomous learning experience \cite{Griol2014,kim2020}. In addition, using an automated system such as a chatbot, can draw teachers’ attention to topics that students struggle to comprehend or need further assistance in understanding.

Given the potential benefits that chatbots could bring to the classroom, the focus of the EDUBOTS KA2 European project \footnote{https://www.edubots.eu}, the sponsor of this study, is to gather, analyse and disseminate best practices on the use of chatbots in higher education institutions.  
% Antonio - Actually there was a previous study focused on students and the present one, focused on teachers. I'll try to explain this, please revise the text below. ;)
With this aim, the authors conducted a previous study aimed to collect feedback from students about their actual needs \cite{MoraChatbots2021}.  Hence, the objective was to collect information about students’ preferences of tools/applications for chatting and messaging, how they use them in educational context, who do they like to be with in the class messaging groups, and their expectations of the chatbots in terms of assisting them during their learning process. Therefore, two surveys for bachelor and master degree students at the University of Granada (Spain) were designed and answers from more than 250 students were collected. The key findings of the students surveys reveal that they prefer to use messaging applications, such as Telegram or WhatsApp as they are familiar to them. Students also expressed interest in chatbots that can assist them with organizing their course schedules, help them access their assignment grades and facilitate searching for course’s resources.

The results of this first phase were fed into the second phase of the study, which is addressed in the present work. It is focused on the other main actor in education process: the teacher. Thus, we have now collected and analised teachers’ opinions about using chatbots and messaging applications in classes.
To this end two new surveys have been developed, and responses from almost 500 teachers (282 + 205) were collected and analysed.  This study focuses on discussing the implications and the impact of the development of future technology-enhanced tools and educational institutions’ policies by finding answers to the following research questions:
\begin{itemize}
\item RQ1 - Are teachers already using messaging apps in their classes? 
\item RQ2 - Which kind of chatbots would teachers find useful in their classes? 
\item RQ3 - Which kind of interaction do teachers prefer with their students?
\item RQ4 - What kind of interaction media features do teachers value the most?
\end{itemize}

@@

\begin{figure}[h!tbp]
\centering
<<messaging_organization, echo=FALSE, fig.width=19/2.54, fig.height=5/2.54>>=
survey.pilot.MessagingOrganization <- within(survey.pilot.MessagingOrganization, Messaging.Organization <- factor(Messaging.Organization, levels=names(sort(table(Messaging.Organization), decreasing=TRUE))))
ggplot(survey.pilot.MessagingOrganization, aes( x = Messaging.Organization, y = ..count.., group=Sector, fill = Sector ) ) + geom_bar()+ theme(axis.text.x = element_text(angle = 90))+coord_flip()+ theme(axis.title.y=element_blank())
@
%\includegraphics[width=\textwidth]{figures/messaging_organization.pdf}
\caption{Distribution of teachers' preferences for the chat groups with their students: from groups only with their students from a specific course to groups with greater social interaction with all students in their School or Faculty.}
\label{figure:chatbots_messagingorganization}
\end{figure}

@@

\begin{table}[h!tbp]
\begin{tabular}{lllll}
\hline
\multicolumn{1}{c}{Kind of interactions} & \multicolumn{2}{c}{Yes} & \multicolumn{2}{l}{No} \\ \cline{2-5} 
\multicolumn{1}{c}{} & \multicolumn{1}{c}{Frequency} & \multicolumn{1}{c}{\%} & Frequency & \multicolumn{1}{c}{\%} \\ \hline
\begin{tabular}[c]{@{}l@{}}Chat interactions among \\ students in the same course\end{tabular} & 171 & 60.6 & 110 & 39.0 \\
\begin{tabular}[c]{@{}l@{}}Chat interactions among students, \\ teachers, and the faculty\end{tabular} & 13 & 4.6 & 269 & 95.4 \\
\begin{tabular}[c]{@{}l@{}}Chat interactions among students \\ of the same study year and teachers\end{tabular} & 51 & 18.1 & 231 & 81.9 \\
\begin{tabular}[c]{@{}l@{}}Teacher not being part of the \\ interaction\end{tabular} & 113 & 40.1 & 169 & 59.9 \\ \hline
\end{tabular}
\vspace{0.8em}
\caption{Kind of interactions preferred}
\label{tab:kindinteractions}
\end{table}

@@

%Olia - Isn't the following paragraph better fitting for RQ3? What kind interactions do teachers prefer with their students? 
% Moved here
Regarding the social factor of chat groups that teachers use in class, according to Fig. \ref{figure:chatbots_messagingorganization} the vast majority prefer small groups only with the students from the same course. These are more focused groups with specific goals and dedicated to the organization of the course and its tasks, and from the pedagogic point of view seem more appropriate to improve the learning process. Interestingly, and as a cross-check of the answers above, about 30\% of teachers consider they should not be part of the chat group.
This might seem to contradict the results of the other survey, but in fact, being a result of different surveys, to a certain point affirms the same thing: there is a great amount of teachers that would be against being in a chat group with students. Whether that's a slim majority or not, that's open to debate (and would probably need a more pointed survey focused on this), but what is clear is that forcing teachers to create chat groups with students and participate in them would create a certain amount of resistance.
Also, only university teachers find interesting a group with all the students and teachers in their own Faculty or School. The lack of teachers from vocational education here may be the consequence of using specific language such as "Faculty". Moreover, the fact that many universities and schools already use these groups for administrative and social interaction (e.g. \cite{dibitonto2018chatbot}) might be the reason for the low percentage of teachers that selected this response.

@@

As it appears in the responses presented above, teachers’ responses are distributed between positive and negative in relation to the media features: analytics, connectivity, and hidden phone, while for the features: familiarity, horizontally, sustainability and officiality the majority provided a positive response, designating their preference to these features. On the other hand, the majority of teachers seems to consider not valuable the presence of official formation, pluggability, unidirectionality and synchrony in the interaction media features of messaging apps.

@@

A chi-square test of independence was performed to examine the potential relationship among the various features. The statistically significant correlations are given in Table \ref{tab:interactionfeatures}.

@@

Finally, a two-step cluster analysis was conducted, with a log-likelihood distance measure adopted, to explore how the cases could be grouped based on their preferences of specific interaction media features, and which features have larger predictor importance for the clustering (see Fig. \ref{figure:predictor}). Two clusters occurred from the analysis: cluster 1 (n=78, 60.9\%) was comprise of instructors who did not value interaction media pluggability, connectivity, official formation, but valued interaction media analytics, familiarity, officiality, while cluster 2 (n=50, 39.1\%) was comprise of instructors from which the majority did not value media analytics, official formation, familiarity, but valued pluggability, connectivity, and officiality. Interaction media pluggability appears to have the most important predictor importance in the clustering of cases (predictor importance = 1.0), whilst media unidirectionality was the least important factor (predictor importance = 0.04). The cluster quality is fair, but not a good one (silhouette measure of cohesion and separation).

>>>>

A \textit{chatbot} is a program, sometimes developed using Artificial Intelligence techniques, able to communicate in a similar way as humans do \cite{Gong2008}, using, in some cases, natural language. Indeed, most of these applications would be, apparently, close to passing a classic Turing test \cite{moor2003turing}, since they are able to answer almost any question fluently, and even pose own questions in their conversations. The recent advancement in this discipline has played an important role in many fields especially in education and online tutoring \cite{clarizia2018,Smutny2020}. These automatic systems facilitate the delivery of personalized learning by adapting to students’ pace of learning and providing customized online tutoring outside the classroom. Thus, chatbots can significantly contribute to providing interactive learning experiences as well as improving individual attention \cite{agarwal2020}.
 In this sense, chatbot technology offers a great opportunity for the improvement of tutoring systems \cite{Daniel2016,agarwal2020}, as not all students are comfortable with face-to-face tutoring with the instructor. In fact, in many cases, some students experience stress because of their need to ask a question in front of the entire class. This is why many students resort to silence  
 % Antonio - I guess refrain to silence is the opposite of what we wanted to say, isn't it? :D
 % mistaking "resort" for "refrain". You don't say "refrain to", but refrain from.
 in the classroom and tend to contact their teachers later via email.
 Consequently, this can lead to not only a delay in obtaining answers to their questions butcan also cause a significant increase in their teachers’ workload, especially when factoring in the disproportion between the class size and the number of teachers. Hence, the chatbot technology has a potential to mitigate this problem by providing answers to students’ questions and facilitating a dynamic and autonomous learning experience \cite{Griol2014,kim2020}. In addition, using an automated system such as a chatbot, can draw teachers’ attention to topics that students struggle to comprehend or need further assistance in understanding.

Given the potential benefits that chatbots could bring to the classroom, the focus of the EDUBOTS KA2 European project \footnote{https://www.edubots.eu}, the sponsor of this study, is to gather, analyse and disseminate best practices on the use of chatbots in higher education institutions.

 The line of research of which this paper is a part has involved two phases of data collection.
 \begin{itemize}
\item The first phase aimed to collect feedback from students about their actual needs \cite{MoraChatbots2021}. Hence, the objective was to collect information about students’ preferences of tools/applications for chatting and messaging, how they use them in educational context, who do they like to be with in the class messaging groups, and their expectations of the chatbots in terms of assisting them during their learning process. Therefore, two surveys for bachelor and master degree students at the University of Granada (Spain) were designed and answers from more than 250 students were collected.   The key findings of the students surveys reveal that they prefer to use messaging applications, such as Telegram or WhatsApp as they are familiar to them. Students also expressed interest in chatbots that can assist them with organizing their course schedules, help them access their assignment grades and facilitate searching for course’s resources. 
\item The results of the first phase were fed into the second phase of the study, which focused on teachers’ opinions about using chatbots and messaging applications in classes. This is what is going to be presented in this paper.
\end{itemize}

Initially, a single survey was developed for this purpose. Responses from 300 teachers were collected and analysed. An initial analysis led to the creation of a second survey, which collected responses from 200 different higher education teachers. The questions on this survey focus on discussing the implications and the impact of the development of future technology-enhanced tools and educational institutions’ policies by finding answers to the following research questions:

\begin{itemize}
\item RQ1 - Are teachers already using messaging apps in their classes? 
\item RQ2 - Which kind of chatbots would teachers find useful in their classes? 
\item RQ3 - Which kind of interaction do teachers prefer with their students?
\item RQ4 - What kind of interaction media features do teachers value the most?
\end{itemize}

@@

\begin{figure}[h!tbp]
\centering
<<messaging_organization, echo=FALSE, fig.width=19/2.54, fig.height=5/2.54>>=
survey.pilot.MessagingOrganization %>%mutate(Messaging.Organization = ifelse(as.character(Messaging.Organization) == "A chat with all the students and teachers in my Faculty/School", "A chat with all the students and teachers in my Faculty/School/all years", as.character(Messaging.Organization))) %>%
  mutate(Messaging.Organization = ifelse(as.character(Messaging.Organization) == "A chat with all the students and teachers in the same studies (all years)", "A chat with all the students and teachers in my Faculty/School/all years", as.character(Messaging.Organization)))->df2

df2 %>%mutate(Messaging.Organization = ifelse(as.character(Messaging.Organization) == "A chat with all the students and teachers in my Faculty/School/all years", "Chat interactions among all students, teachers of the School/Faculty", as.character(Messaging.Organization))) %>%
  mutate(Messaging.Organization = ifelse(as.character(Messaging.Organization) == "I'd rather leave students self-organize and not be part of the chat", "Teacher not being part of the interaction", as.character(Messaging.Organization)))%>%
  mutate(Messaging.Organization = ifelse(as.character(Messaging.Organization) == "A chat with my students in the same course", "Chat interactions among students in the same course", as.character(Messaging.Organization)))%>%
  mutate(Messaging.Organization = ifelse(as.character(Messaging.Organization) == "A chat with all students and teachers in the same semester/year", "Chat interactions among students of the same study year and teachers", as.character(Messaging.Organization)))->df2

df2 <- within(df2, Messaging.Organization <- factor(Messaging.Organization, levels=names(sort(table(Messaging.Organization), decreasing=TRUE))))
ggplot(df2, aes( x = Messaging.Organization, y = ..count.., group=Sector, fill = Sector ) ) + geom_bar()+ theme(axis.text.x = element_text(angle = 90))+coord_flip()+ theme(axis.title.y=element_blank())
@
%\includegraphics[width=\textwidth]{figures/messaging_organization.pdf}
\caption{Distribution of teachers' preferences for the chat groups with their students: from groups only with their students from a specific course to groups with greater social interaction with all students in their School or Faculty.}
\label{figure:chatbots_messagingorganization}
\end{figure}

@@

\begin{table}[h!tbp]
\begin{tabular}{lllll}
\hline
\multicolumn{1}{c}{Kind of interactions} & \multicolumn{2}{c}{Yes} & \multicolumn{2}{l}{No} \\ \cline{2-5} 
\multicolumn{1}{c}{} & \multicolumn{1}{c}{Frequency} & \multicolumn{1}{c}{\%} & Frequency & \multicolumn{1}{c}{\%} \\ \hline
%\begin{tabular}[c]{@{}l@{}}Chat interactions among \\ students in the same course\end{tabular} & 171 & 60.6 & 110 & 39.0 \\
%\begin{tabular}[c]{@{}l@{}}Chat interactions among students, \\ teachers of the School/Faculty\end{tabular} & 13 & 4.6 & 269 & 95.4 \\
%\begin{tabular}[c]{@{}l@{}}Chat interactions among students \\ of the same study year and teachers\end{tabular} & 51 & 18.1 & 231 & 81.9 \\
%\begin{tabular}[c]{@{}l@{}}Teacher not being part of the \\ interaction\end{tabular} & 113 & 40.1 & 169 & 59.9 \\ \hline
\begin{tabular}[c]{@{}l@{}}Chat interactions among \\ students in the same course\end{tabular} & 166 & 59.1 & 115& 40.9 \\
\begin{tabular}[c]{@{}l@{}}Chat interactions among students, \\ teachers of the School/Faculty\end{tabular} & 19 & 6.7 & 266 & 94.3 \\
\begin{tabular}[c]{@{}l@{}}Chat interactions among students \\ of the same study year and teachers\end{tabular} & 58 & 20.5 & 224 & 79.4 \\
\begin{tabular}[c]{@{}l@{}}Teacher not being part of the \\ interaction\end{tabular} & 81 & 28.7 & 201 & 71.3 \\ \hline
\end{tabular}
\vspace{0.8em}
\caption{Kind of interactions preferred}
\label{tab:kindinteractions}
\end{table}

@@

Regarding the social factor of chat groups that teachers use in class, according to Fig. \ref{figure:chatbots_messagingorganization} the vast majority prefer small groups only with the students from the same course. These are more focused groups with specific goals and dedicated to the organization of the course and its tasks, and from the pedagogic point of view seem more appropriate to improve the learning process. Interestingly, and as a cross-check of the answers above, about 30\% of teachers consider they should not be part of the chat group.
This might seem to contradict the results of the other survey, but in fact, being a result of different surveys, to a certain point affirms the same thing: there is a great amount of teachers that would be against being in a chat group with students. Whether that's a slim majority or not, that's open to debate (and would probably need a more pointed survey focused on this), but what is clear is that forcing teachers to create chat groups with students and participate in them would create a certain amount of resistance.
Also, only university teachers find interesting a group with all the students and teachers in their own Faculty or School. The lack of teachers from vocational education here may be the consequence of using specific language such as "Faculty". Moreover, the fact that many universities and schools already use these groups for administrative and social interaction (e.g. \cite{dibitonto2018chatbot}) might be the reason for the low percentage of teachers that selected this response.

@@

As it appears in the responses presented above, teachers’ responses are distributed between positive and negative in relation to the media features: analytics, connectivity, and hidden phone, while for the features: familiarity, horizontally, sustainability and officiality the majority provided a positive response, designating their preference to these features. On the other hand, the majority of teachers seems to consider not valuable the presence of official formation, pluggability, unidirectionality and synchronization in the interaction media features of messaging apps. In other words, the majority of the teachers seem to prefer to use messaging apps to which they are already familiar, and which have been granted by their academic institution, normally also accompanied with technical support by the university’s staff and which do not require any actions on the behalf of the teachers for maintenance. In addition, the majority of the teachers value more the messaging apps in which there is no hierarchy in the communication level among the different users (i.e., teachers, students), and when the use of the app does not require or reveal their personal phone number. Last, the majority of the teachers value the most messaging apps that provide analytics in a form of a dashboard.

@@

A chi-square test of independence was performed to examine the potential relationship among the various features. The statistically significant correlations are given in Table \ref{tab:interactionfeatures}. As indicated from the chi square results, the presence of the analytics feature in messaging apps, has been found to be positively correlated with the presence of features such as, connectivity, familiarity, official formation, pluggability and sustainability. This means, that teachers who prefer to use an app (or tool) that provides analytics about students’ actions and/or performance, most probably also prefer this app to be able to be connected to other applications, offer different plugins to the teachers and be an app with which the teachers are already familiar, while receiving some training from the university about its use and do not have to worry about its maintenance. In addition, the hidden phone feature was found to be correlated with the features: official formation (i.e., the university offers some training courses or talks about the tool/app) and unidirectionality (i.e., only the teacher can publish information in the channel). This can be expected, as teachers might prefer not to reveal personal data, such as their personal phone number, when the tool is being introduced by the university and when the level of communication is one-way (i.e., from the teacher to the students). Furthermore, pluggability has been found to be correlated with official formation, implying that teachers who prefer pluggability as a feature in a messaging app/tool, they should also prefer to having this tool/app being properly and officially introduced by the university and offer some training about its usage.

@@

Finally, a two-step cluster analysis was conducted, with a log-likelihood distance measure adopted, to explore how the cases could be grouped based on their preferences of specific interaction media features, and which features have larger predictor importance for the clustering (see Fig. \ref{figure:predictor}). Two clusters occurred from the analysis: cluster 1 (n=78, 60.9\%) was comprise of instructors who did not value interaction media pluggability, connectivity, official formation, but valued interaction media analytics, familiarity, officiality, while cluster 2 (n=50, 39.1\%) was comprise of instructors from which the majority did not value media analytics, official formation, familiarity, but valued pluggability, connectivity, and officiality. The formation of the two clusters demonstrates different trends that appear to exist in teachers’ responses, based on their most preferred interaction media features. In other words, there is a group of teachers who value the most certain features, such as being familiar with an application, having access to analytics, and having the support from the university when using the application, while a second group of teachers seems to value mostly the features of pluggability, the capability of the app to connect with other apps, and receiving training from the university before/while using the app for teaching purposes. Interaction media pluggability appears to have the most important predictor importance in the clustering of cases (predictor importance = 1.0), whilst media unidirectionality was the least important factor (predictor importance = 0.04). The cluster quality is fair, but not a good one (silhouette measure of cohesion and separation).
